---
title: "Lab 2 part 4"
author: "Carlos Schrupp"
format: pdf
---

Loading libraries

```{r}
#| echo: false
library(haven)
library(here)
library(lmtest)
library(sandwich)
library(tidyverse)
library (survey)
```

Loading part2 data

```{r}
#| echo: false
part2 = read.csv("C://Users//Carlos//Downloads//part2.csv")
head(part2)
```

Adding age groups:

```{r}
#| echo: true
part2$age_group <- cut(part2$age, 
                       breaks = c(0, 10, 20, 30, 40, 50, 60, 70, Inf), 
                       labels = c("0-10","10-19","20-29", "30-39", "40-49", "50-59", "60-69", "70+"), 
                       right = FALSE)
```

Adding income categories:

```{r}
part2$income_category <- cut(part2$income,
                             breaks = c(-Inf, 1.3, 3.0, Inf),
                             labels = c("Low", "Middle", "High"),
                             right = FALSE)
```

Adding obesity flag:

```{r}
# Assuming your dataset 'part2' has a numeric 'bmi' variable:
part2$bmi_category <- cut(part2$bmi, 
                          breaks = c(0, 18.5, 24.9, 29.9, Inf),
                          labels = c("Underweight", "Normal weight", "Overweight", "Obesity"))

summary(part2$bmi_category)

# Create an obesity flag
part2$obese_flag <- ifelse(part2$bmi >= 30, 1, 0)
```

Filtering age \<20

```{r}
part2 <- part2 %>% 
  filter(age >= 20)
```

Set ethnicity reference to "White"

```{r}
part2$ethnicity <- factor(part2$ethnicity, 
                          levels = c("Asian", "White", "Black", "Mexican", "Other Hispanic", "Other Race"))
part2$ethnicity <- relevel(part2$ethnicity, ref = "White")
levels(part2$ethnicity)
```

Filter Asian ethnicity:

```{r}
part2 <- subset(part2, ethnicity != "Asian")
```

Load NHANES survey stratification and weights:

```{r}
#| echo: true

nhanes_survey = nhanes_design <- svydesign(
  id = ~psu,
  strata = ~stratum,
  weights = ~weight,
  data = part2,
  nest = TRUE
)
```

Calculating means to know where we stand:

Note: Age has a complex relationship with BMI, its recommended to start using BMI with ages 20 and up

```{r}
#| echo: true
svymean(~bmi, design = nhanes_design)    # Weighted mean of BMI


svyby(~bmi, ~cut(income, breaks=c(0,1,2,3,4,5)), nhanes_design, svymean) # Means by income groups


svyby(~bmi, ~cut(part2$age, breaks=c(0,2,6,12,20,Inf)), nhanes_design, svymean) # Means by age groups

# svyby(~income, ~cut(part2$ethnicity, breaks=c("Asian")), nhanes_design, svymean)

```

Model 1

Regressing income alone on BMI

Result it's not significant with income alone

```{r}

model_1 <- svyglm(bmi ~ income  , design = nhanes_survey)
summary(model_1)
```

```{r}



model_1a <- svyglm(bmi~ income + ethnicity , design = nhanes_survey)

summary(model_1a)
```

```{r}
# Extract fitted values and residuals from the model
fitted_vals <- fitted(model_1a)
residual_vals <- residuals(model_1a)

# Put them into a data frame for plotting
resid_data <- data.frame(
  fitted = fitted_vals,
  residuals = residual_vals
)

library(ggplot2)

ggplot(resid_data, aes(x = fitted, y = residuals)) +
  geom_point(alpha = 0.5) +         # Scatter plot of residuals
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") + 
  labs(title = "Residuals vs. Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```

```{r}
ggplot(part2, aes(x = income, y = bmi)) +
  geom_point(alpha = 0.5) +
  labs(title = "BMI vs. Income",
       x = "Income",
       y = "BMI") +
  theme_minimal()
```

```{r}
# Summarize data to find proportion of obesity within each income category
income_obesity_summary <- part2 %>%
  group_by(income_category) %>%
  summarize(obesity_rate = mean(obese_flag, na.rm = TRUE))

# Plot the obesity rate by income category
ggplot(income_obesity_summary, aes(x = income_category, y = obesity_rate)) +
  geom_col(fill = "steelblue") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Obesity Prevalence by Income Category",
    x = "Income Category",
    y = "Obesity Rate (%)"
  ) +
  theme_minimal()

```

Model 2

Regressing income and age on BMI

Result: Using age and income results on significance for both

```{r}
model_2 <- svyglm(bmi ~ income + age, design = nhanes_survey)
summary(model_2)
```

Model 3

Regressing income, age and a factored etnicity on BMI

Result: Using age, ethnicity and income results on significance for all, but seems to reduce it on income

```{r}

model_3 <- svyglm(bmi ~ income + age + factor(ethnicity), design = nhanes_survey)
summary(model_3)

```

Model 4

Regressing income, factored age and factored ethnicity on BMI

Result: Income is not significant any more

```{r}
model_4 <- svyglm(bmi ~ income + factor(age_group) + factor(ethnicity), 
                       design = nhanes_survey)
summary(model_4)
```

```{r}
model_5 <- svyglm(income ~ ethnicity, 
                       design = nhanes_survey)
summary(model_5)
```

```{r}

```

Questions:

What exactly are the weighting and sampling of nhanes doing with the data?.

What kind of model are we really applying, I would like to know more details on "svyglm"

References:

*U.S. Department of Health and Human Services, Centers for Disease Control and Prevention (CDC), National Center for Health Statistics. (2021). NHANES Analytic and Reporting Guidelines.* Retrieved from: <https://wwwn.cdc.gov/nchs/nhanes/analyticguidelines.aspx>

World Health Organization. (2000). *Obesity: Preventing and Managing the Global Epidemic.* WHO Technical Report Series 894. <http://www.who.int/nutrition/publications/obesity/WHO_TRS_894/en/>
