---
title:  "Lab 2: Great American Paradox"
subtitle: "BMI relationship to household income in the U.S. using the National Health and Nutrition Examination Survey (NHANES), collected between August 2021 and August 2023"
date: today
author: "Datasci 203 Team 2 - G. Frimpong, K. Coppa, R. Schaefer, C. Schrupp"

abstract: This study investigates the relationship between income and Body Mass Index (BMI) to explore the "Great American Paradox," where economic prosperity coexists with high rates of obesity and health disparities. Using data from the National Health and Nutrition Examination Survey (NHANES) collected between 2021 and 2023, we analyze a nationally representative sample of 5,194 individuals. The dataset includes directly measured BMI and socioeconomic variables, ensuring robust and reliable analysis. Regression models evaluate how income-to-poverty ratios influence BMI while accounting for demographic factors such as age, gender, and race/ethnicity. Results highlight disparities in health outcomes across income levels, providing insights into the socioeconomic determinants of obesity. This research informs public health strategies aimed at reducing obesity-related disparities in the United States.
toc: false 
colorlinks: true
cap-location: bottom
execute: 
  echo: false 
  warning: false 
  message: false 
bibliography: references.bib
citeproc: true
biblio-title: References
link-citations: true
format: 
  pdf: 
    documentclass: scrreprt
---

```{r}
### Load libraries
library(haven)
library(here)
library(lmtest)
library(sandwich)
library(tidyverse)
library(survey)
library(car)
library(ggplot2)
library(stargazer)
library(gridExtra)
library(viridis)
```

```{r}
### Set theme
theme_set(theme_minimal())
```

**INTRODUCTION**

Obesity is often viewed as a consequence of individual lifestyle choices, yet in the United States, it starkly reveals systemic inequalities—a phenomenon known as the "Great American Paradox." This paradox, where economic prosperity coexists with high rates of obesity and related health disparities, challenges assumptions about wealth as a protective factor for health.
Surprisingly, obesity disproportionately affects lower-income groups, raising critical questions about how socioeconomic conditions interact with health behaviors and outcomes.

This study examines the relationship between income and Body Mass Index (BMI) to uncover the socioeconomic mechanisms underpinning this paradox.
By leveraging robust, nationally representative data, we aim to identify patterns and disparities that contribute to obesity's persistence despite economic growth.
The results will illuminate the broader social determinants of health, offering evidence to shape targeted public health interventions.
In doing so, this research seeks to provide actionable insights into addressing one of America’s most pressing public health challenges.

**DESCRIPTION OF THE DATA SOURCE**

To analyze the relationship between income and BMI, we utilize data from the National Health and Nutrition Examination Survey (NHANES), collected between August 2021 and August 2023.
NHANES is a cornerstone of public health research in the United States, known for its rigorous methodology combining interviews and standardized physical examinations conducted by trained professionals.
This ensures reliable, directly measured health data, particularly critical for BMI analysis.

```{r}
### Load data

# Demographics File (DEMO_L)
nhanes_demo = read_xpt(url("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DEMO_L.xpt"))

# Body Measurements File (BMX_L)
nhanes_bm = read_xpt(url("https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/BMX_L.xpt"))
```

**DATA WRANGLING**

Before we modified the data, the demographics file included 11,933 observations and 27 variables and the body measurements files included 8,860 observations and 22 variables.
Both of these datasets include a unique ID that we used to join them.
The joined dataset had 8,860 observations and 48 variables.
We then made two modifications to the joined dataset to prepare it for analysis.
The first was renaming the relevant columns to be more human readable.
While this is not a required preprocessing step, the NHANES column names are not always clear, so we found it easier to work with it if we renamed the columns.
The second modification was updating the values in the categorical features.
NHANES encodes all categorical variables with numeric indicators, so we replaced them with descriptive labels to make the data more human readable and to ensure the regression model treats these features as categorical variables.
The ethnicity categorical variable, reference was set to "White" to ensure stability in the regression model as its the largest group.

```{r}
### Data Wrangling 

# Join datasets on unique ID
nhanes_df = nhanes_demo %>%
  inner_join(nhanes_bm, by = "SEQN") 

nhanes_df = nhanes_df %>%
  # Select wanted columns
  select(SDMVSTRA, SDMVPSU, WTMEC2YR, BMXBMI, INDFMPIR, RIDAGEYR, RIAGENDR, RIDRETH3) %>%
  # Make names more readable
  rename(
    stratum = SDMVSTRA,
    psu = SDMVPSU,
    weight = WTMEC2YR,
    bmi = BMXBMI,
    income = INDFMPIR,
    age = RIDAGEYR,
    gender = RIAGENDR,
    ethnicity = RIDRETH3
  ) %>%
  # Add labels to categorical levels
  mutate(
    gender = case_when(
      gender == 1 ~ "Male",
      gender == 2 ~ "Female"
    ),
    ethnicity = case_when(
      ethnicity == 1 ~ "Mexican",
      ethnicity == 2 ~ "Other Hispanic",
      ethnicity == 3 ~ "White",
      ethnicity == 4 ~ "Black",
      ethnicity == 6 ~ "Asian",
      ethnicity == 7 ~ "Other Race"
    )
  )

# Set reference ethnicity to "White"
nhanes_df$ethnicity <- factor(nhanes_df$ethnicity, 
                          levels = c("Asian", 
                                     "White", 
                                     "Black", 
                                     "Mexican", 
                                     "Other Hispanic",
                                     "Other Race"))
nhanes_df$ethnicity <- relevel(nhanes_df$ethnicity, ref = "White")
```

**OPERATIONALIZATION**

Out of the 48 variables in the joined dataset, we kept 8 of them for our analysis.
Three of them are used for stratifying the data.
NHANES uses a complex stratified sampling method to ensure the data is representative of the American population.
The stratification variables are used to ensure that our models match the stratification methods used by NHANES.
Our dependent variable is BMI and our main independent variable is income, represented as a ratio of a person’s income over the poverty line.
We also kept the age, gender, and ethnicity variables as we hypothesize these demographic factors may also be significant predictors of BMI.
Out of the 8,860 observations in the joined dataset, 1,480 are missing data in at least one of the kept variables.
We dropped these observations to ensure we could conduct our full analysis with all of the remaining data.
We also dropped 2,186 observations where the subject’s age is less than 20 years old.
We found that BMI was random noise for people under 20, so we removed these observations to get more reliable results on the remaining data @Johnson2013.
After selecting variables and dropping observations, our final dataset for analysis contained 5,194 observations and 8 variables.
An income categorical variable is created to facilitate the examination of the nonlinear relationship between income and BMI, with a cutoff of 1.3 chosen to distinguish low-income groups at or near the poverty threshold, as defined by the income-to-poverty ratio @Drewnowski2009.

```{r}
### Operationalization

# Remove rows with any NA's
nhanes_df = nhanes_df %>%
  drop_na()

# Drop rows where age is less than 20
nhanes_df = nhanes_df %>%
  filter(age >= 20)

# Create income categorical variable
nhanes_df$income_category <- cut(nhanes_df$income,
                             breaks = c(-Inf, 1.3, Inf),
                             labels = c("Low", "High"),
                             right = FALSE)

# Create obesity flag based on bmi
nhanes_df$obese_flag <- ifelse(nhanes_df$bmi >= 30, 1, 0)
nhanes_df$obese_flag = as.factor(nhanes_df$obese_flag)

# Create NHANES survey stratification and weights model
nhanes_survey <- svydesign(
  id = ~psu,
  strata = ~stratum,
  weights = ~weight,
  data = nhanes_df,
  nest = TRUE
)
```

**DATA VISUALIZATION**

```{r}
## Density plots of Income and BMI (with obesity threshold) 
# density plot - income
inc.dens.plot = ggplot(nhanes_df, aes(x = income)) +
  geom_density(fill = "#0f39ab", alpha = 0.7) +  # Blue fill with transparency
  labs(
    title = "Density Plot of Income",
    x = "\nIncome",
    y = "Density\n"
  ) +
  theme_minimal() +
  theme(
    axis.line = element_line(),
    panel.grid = element_blank(),
    axis.text = element_text(size = 14), 
    axis.title = element_text(size = 16),
    plot.title = element_text(size = 20, hjust = 0.5)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))
# density plot - bmi
bmi.dens.plot = ggplot(nhanes_df, aes(x = bmi)) +
  geom_density(fill = "#0f39ab", alpha = 0.7) +
  geom_vline(xintercept = 30, linetype = "dashed", color = "gray", size = 1) +
  labs(
    title = "Density Plot of BMI",
    x = "\nBMI",
    # remove y label as the plots will be side by side
    y = "\n"
  ) +
  theme_minimal() +
  theme(
    axis.line = element_line(),
    panel.grid = element_blank(),
    axis.text = element_text(size = 14), 
    axis.title = element_text(size = 16),
    plot.title = element_text(size = 20, hjust = 0.5)
  ) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))

## Arrange plots in 2 columns - horizontal
grid.arrange(
  inc.dens.plot, 
  bmi.dens.plot, 
  ncol = 2
)
```
```{r}
# Obesity Rates by Income/Ethnicity sub-groups
inc.eth.out = nhanes_df %>%
  group_by(ethnicity, income_category) %>%
  summarise(n=n(),
            perc_obs = round(sum(obese_flag==1)/n*100,1))

inc.eth.plot = ggplot(inc.eth.out, aes(x = income_category, y = reorder(ethnicity,n), fill = perc_obs)) +
  geom_tile(width=0.95, height=0.95) +
  scale_fill_viridis(
    option = "plasma",  # "magma", "plasma", "inferno", "viridis"
    direction = -1
  ) +
  labs(x = "\nIncome Group", 
       y = "Ethnicity\n", 
       fill = "Obesity %", 
       title = "How Obesity Varies by Income Group and Ethnicity"
       ) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text = element_text(size=14),
        text = element_text(size=16))
```

Figure 1: Density plots of the income and bmi variables

```{r fig-cap="How Obesity Varies by Income Group and Ethnicity"}
inc.eth.plot
```

Figure 2: How Obesity Varies by Income Group and Ethnicity
This heatmap shows how obesity prevalence varies by income and ethnicity. Asians are outliers with an extremely low prevalence of obesity (low-6.7%, high-11.7%) compared to the other ethnic groups. Black, Mexican, and Other Hispanic ethnicities have a higher prevalence of obesity in the high-income group.


```{r}
bp.df = nhanes_df %>%
  group_by(income_category, gender) %>%
  summarise(n=n(),
            perc_obs = round(sum(obese_flag==1)/n*100,1))

inc.gnd.plot = ggplot(bp.df, aes(x = income_category, y = perc_obs, fill = gender)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(
    values = c("Male" = "#31688EFF", "Female" = "#35B779FF")  
  ) +
  labs(
    x = "\nIncome Group",
    y = "Obesity Percentage\n",
    title = "Obesity Prevalence by Income Group and Gender",
    fill = "Gender"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size=14),
    text = element_text(size=16)
  )
```

```{r fig-cap="Obesity Prevalence by Income Group and Gender"}
inc.gnd.plot
```

Figure 3: Obesity Prevalence by Income Group and Gender
This figure shows that females have a higher prevalence of obesity compared to males. Females who have low income also have a higher obesity rate than females with a high income. This relationship is inverted for males.


**MODEL SPECIFICATION**

1.  **Model Progression:**

    -   **Model 1:** Baseline model with income as a continuous predictor.

    -   **Model 1a:** Uses income categories instead of continuous income.

    -   **Model 2:** Adds gender and its interaction with income.

    -   **Model 3:** Includes age as a predictor and its interaction with income.

    -   **Model 4:** Adds ethnicity and its interaction with income.

#### **Model 1: Income Alone**

```{r}
model_1 <- svyglm(bmi ~ income  , design = nhanes_survey)
summary(model_1)
```

-   **Income (-0.33, p \< 0.01):**

    -   A significant negative association between income and BMI.
        For every unit increase in income, BMI decreases by approximately 0.33 units.

    -   Suggests that individuals with higher income tend to have lower BMI, on average.

#### **Model 1a: Income Categories**

```{r}
model_1a <- svyglm(bmi ~ income_category  , design = nhanes_survey)
summary(model_1a)
```

-   **Income Category: High (-0.81, p \< 0.05):**

    -   Individuals in the "High" income category have a BMI approximately 0.81 units lower than those in the reference income category ("Low").

    -   The relationship remains significant but provides a categorical perspective on the income-BMI association.

#### **Model 2: Adding Gender.**

```{r}
model_2 <- svyglm(bmi~ income * gender , design = nhanes_survey)
summary(model_2)
```

-   **Income (-0.64, p \< 0.01):**

    -   Similar to Model 1, income remains negatively associated with BMI.

-   **Gender: Male (-2.60, p \< 0.01):**

    -   Males have a significantly lower baseline BMI (by 2.60 units) compared to females.

-   **Income × Gender (0.67, p \< 0.01):**

    -   Interaction term indicates that the negative income-BMI relationship is less pronounced for males. For males, the slope of the income-BMI relationship is approximately (-0.64 + 0.67 = +0.03), suggesting a negligible or slightly positive association.

#### **Model 3: Adding Age.**

```{r}
model_3 <- svyglm(bmi ~ income * age, design = nhanes_survey)
summary(model_3)
vif(model_3)
```

-   **Income (-0.47, not significant):**

    -   The negative income-BMI relationship weakens and becomes statistically insignificant when age is added.

-   **Age (0.67, p \< 0.01):**

    -   Positive association: BMI increases slightly with age, likely reflecting physiological changes or age-related trends in weight.

-   **Income × Age (not significant):**

    -   No evidence that the relationship between income and BMI varies systematically with age.

#### **Model 4: Adding Ethnicity.**

```{r}
model_4 <- svyglm(bmi~ income * ethnicity , design = nhanes_survey)
summary(model_4)

vif(model_4)
```

-   **Income (-0.30, not significant):**

    -   The income-BMI relationship remains weak and non-significant in this model.

-   **Ethnicity: Asian (-4.63, p \< 0.01):**

    -   Asian participants have significantly lower BMI (by \~4.63 units) compared to the reference group at the baseline income level.

-   **Other Ethnicity Effects (not significant):**

    -   Differences for other ethnic groups are not statistically significant.

-   **Interaction Terms (not significant):**

    -   The slopes of income-BMI relationships do not differ meaningfully by ethnicity.

#### **Model 5: Three-way interaction: Income, Gender, Ethnicity.**

```{r}
#| warning: false
model_5 <- svyglm(bmi ~ income * gender * ethnicity, design = nhanes_survey)
summary(model_5)
vif(model_5)
```

### **Key Statistics**

1.  **Constant (Intercept):**

    -   Represents the predicted BMI for the reference group at baseline levels of the predictors.

    -   For example, in Model 1, the baseline BMI is approximately 30.75 for individuals with zero income.

2.  **Observations:**

    -   All models include 5,194 observations after filtering out individuals under 20 years old.

3.  **Log Likelihood and Akaike Information Criterion (AIC):**

    -   Log likelihood measures model fit; higher (less negative) values indicate better fit.

    -   AIC compares model performance, with lower values indicating a better trade-off between model complexity and fit.

### **Overall Summary**

1.  **Key Findings:**

    -   **Income:** A modest but statistically significant negative relationship with BMI emerges when modeled alone or as categories.

    -   **Gender:** Gender modifies the income-BMI relationship, with males showing a weaker association.

    -   **Age:** BMI increases with age, but the income-BMI relationship does not vary by age.

    -   **Ethnicity:** Significant baseline BMI differences exist (e.g., lower BMI for Asians), but the income-BMI slope does not differ substantially across ethnic groups.

2.  **Model Fit:**

    -   The inclusion of gender (Model 2) improves model fit (lower AIC), while adding age and ethnicity provides additional context but does not drastically improve explanatory power.

3.  **Future Directions:**

    -   Addressing top-coded income values using imputation or external data could refine estimates.

    -   Exploring alternative measures of adiposity (e.g., waist circumference) might yield different insights.

**MODEL ASSUMPTIONS**

**MODEL RESULTS AND INTERPRETATION**

**APPENDIX**

**APPENDIX A - DATA SOURCE**

1.  Demographics File (DEMO_L): Provides our key independent variable: ratio of family income to poverty (INDFMPIR), Includes essential demographic controls: age, gender, race/ethnicity, Contains necessary survey design variables (weights, strata, PSUs) for proper statistical inference, Offers contextual socioeconomic information crucial for understanding income patterns.

    Link: <https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/DEMO_L.xpt>

2.  Body Measurements File (BMX_L): Contains our dependent variable: Body Mass Index (BMI), Includes height and weight measurements taken by trained health technicians, Ensures standardized measurement protocols across all participants, Provides the physical data necessary to explore our central research question.

    Link: <https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2021/DataFiles/BMX_L.xpt>

**APPENDIX B - STATISTICAL MODEL COMPARISON**

```{r}
#| warning: false
# Create stargazer table with the 4 calculated models
stargazer(model_1, model_1a, model_2, model_3, model_4,
          type = "text",   # Options: "text", "html", or "latex"
          title = "Comparison of Models",
          dep.var.labels = c("BMI"),
          column.labels = c("Model 1", 
                            "Model 1a", 
                            "Model 2", 
                            "Model 3",
                            "Model 4"),
          covariate.labels = c("Income", 
                               "Income Category: High", 
                               "Gender: Male", 
                               "Age", 
                               "Ethnicity: Asian", 
                               "Income x Gender", 
                               "Income x Age", 
                               "Income x Ethnicity"),
          omit.stat = c("ll", "aic"),  # Removes Log Likelihood and Akaike stats
          digits = 2)

```

```{r}

```

**APPENDIX C - RESIDUALS VS FITTED VALUES PLOT**

**REFERENCES**

::: {#refs}
:::
